---
title: "Analysis of Mobilizations in El Salvador"
subtitle: "Focus on women's mobilizations and other protests (ACLED)"
author: "Automated Analysis"
format: 
  html:
    theme: litera
    toc: true
    toc-title: "Contents"
    code-fold: true
    code-summary: "Show R code"
execute:
  warning: false
  message: false
---

## 1. Introduction

This document processes and visualizes conflict and protest data in **El Salvador**, using the ACLED database as the primary source. The main focus is to distinguish **women-related mobilizations** from other general protests and riots in the country.

```{r}
#| label: setup

# Load necessary libraries
library(tidyverse)
library(lubridate)
library(ggplot2)

# Configure file path
# Path originally provided with the ACLED database
data_path <- "E:/GFW_New_Analysis/AllCountries2025-6.csv"

# Load the database
# We use read_csv as it is fast and efficient
acled_data <- read_csv(data_path)

# Filter by El Salvador and process dates
sv_data <- acled_data %>% 
  filter(country == "El Salvador") %>%
  mutate(
    # ACLED normally brings an 'event_date' in dmy or ymd format, we handle both common cases
    event_date = as_date(parse_date_time(event_date, orders = c("dmy", "ymd")))
  )
```

## 2. Filtering Methodology

To find the mobilizations, we created a classification by searching for key terms associated with women in the notes and actor columns. If it is not related to women, we check if it is a regular protest or riot.

```{r filtering}
# Keywords for women's mobilizations
keywords <- c(
  "woman", "women", "female", "girl", "girls",
  "gender", "feminism", "feminist",
  "pregnancy", "abortion",
  "rape", "sexual violence", "domestic violence",
  "gender-based violence", "gbv",
  "femicide", "women's rights",
  "mujeres", "mujer", "niñas", "feminista"
)
pattern <- paste(keywords, collapse = "|")

sv_data <- sv_data %>%
  mutate(
    # Search for patterns in lowercase across different columns
    is_women = str_detect(tolower(notes), pattern) |
               str_detect(tolower(actor1), pattern) |
               str_detect(tolower(assoc_actor_1), pattern) |
               str_detect(tolower(actor2), pattern) |
               str_detect(tolower(assoc_actor_2), pattern),
    
    # Fill null values
    is_women = replace_na(is_women, FALSE),
    
    # Create final category
    mobilization_type = case_when(
      is_women ~ "Women's Mobilizations",
      event_type %in% c("Protests", "Riots") ~ "Other Protests/Riots",
      TRUE ~ "Other ACLED Events"
    )
  )

# Keep only the mobilizations/protests for the main analysis
protests_data <- sv_data %>% 
  filter(mobilization_type %in% c("Women's Mobilizations", "Other Protests/Riots"))

cat("Total Women's events:", sum(protests_data$mobilization_type == "Women's Mobilizations"), "\n")
cat("Total Other Protests:", sum(protests_data$mobilization_type == "Other Protests/Riots"), "\n")
```

## 3. Visualizations and Results

### 3.1. Temporal Evolution

We observe how protests have behaved over time to identify mobilization peaks.

```{r timeseries, fig.width=9, fig.height=5}
#| fig-cap: "Evolution of mobilizations over time"

protests_data %>%
  mutate(month_year = floor_date(event_date, "month")) %>%
  count(month_year, mobilization_type) %>%
  ggplot(aes(x = month_year, y = n, color = mobilization_type, group = mobilization_type)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Women's Mobilizations" = "#9b59b6", "Other Protests/Riots" = "#34495e")) +
  theme_minimal() +
  labs(
    title = "Monthly Frequency of Protests in El Salvador",
    subtitle = "Comparison between women's mobilizations and other protests",
    x = "Date",
    y = "Number of Events",
    color = "Event Type"
  ) +
  theme(legend.position = "bottom",
        text = element_text(size = 12))
```

### 3.2. Entities Mentioned in Notes

We extract capitalized phrases from the `notes` column to identify key entities, organizations, or actors mentioned in a narrative way.

```{r entities, fig.width=10, fig.height=6}
#| fig-cap: "Top Entities Extracted from Notes"

library(stringr)

protests_data %>%
  mutate(
    # Extract sequences of 2 or more capitalized words as an approximation to entities
    entities = str_extract_all(notes, "\\b[A-Z][a-z]+(?:\\s+[A-Z][a-z]+)+\\b")
  ) %>%
  unnest(entities) %>%
  # Clean false positives (generic English terms that usually start sentences in ACLED)
  filter(!entities %in% c("El Salvador", "San Salvador", "In El", "At Least", "The National", "Members Of", "A Group", "Police Forces", "The Fmln", "National Civil", "Civil Police", "Armed Forces", "National Civil Police")) %>%
  count(entities, mobilization_type, sort = TRUE) %>%
  group_by(mobilization_type) %>%
  top_n(10, n) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(entities, n), y = n, fill = mobilization_type)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~mobilization_type, scales = "free_y") +
  scale_fill_manual(values = c("Women's Mobilizations" = "#9b59b6", "Other Protests/Riots" = "#34495e")) +
  theme_minimal() +
  labs(
    title = "Top 10 Entities Mentioned in Notes",
    x = "Extracted Entity",
    y = "Mention Frequency"
  )
```

### 3.3. Basic Geographical Distribution

Using the coordinates provided in ACLED, we can map the location of the incidents.

```{r map, fig.width=8, fig.height=6}
#| fig-cap: "Geographical distribution of events"

# Dot plot using latitude and longitude
ggplot(protests_data, aes(x = longitude, y = latitude, color = mobilization_type)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = c("Women's Mobilizations" = "#9b59b6", "Other Protests/Riots" = "#34495e")) +
  theme_minimal() +
  labs(
    title = "Geographical Distribution of Mobilizations",
    subtitle = "El Salvador",
    x = "Longitude",
    y = "Latitude",
    color = "Type"
  ) +
  theme(legend.position = "bottom")
```

## 4. Conclusions

This quick exploratory analysis allows us to observe that mobilization dynamics in El Salvador have distinct peaks depending on whether the agenda is related to women's issues or general conflicts. The organizations involved (actors) also vary significantly.
